---
title: "Optimizing Financial Descions for Corporate Training"
author: "Jaleta Fanta Tesgera"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = 'Test Cases for Mini-Project-Test-Cases/Test Case 1')
library(readxl)
library(lpSolve)
```

## Reading in Files and Building the Matrices
   - - - - - - - - - - - - - - - - - - - 

- Let's read in the three project files (programConflicts.xlsx, skillsNeeded.xlsx, and skillsOffered.xlsx) one at a time build the sets needed for the constraints. 

1. Let's start with skillsNeeded:

# Declare all the  necessary variabless
```{r}
skillsNeeded <- as.matrix(read_excel("skillsNeeded.xlsx"))
nSkills <- ncol(skillsNeeded)-1
nPeople <- nrow(skillsNeeded)
namesOfPeople <- skillsNeeded[,1]
namesOfSkills <- colnames(skillsNeeded)[2:(nSkills+1)]
namesOfSkills
namesOfPeople
nPeople
nSkills
```

# Build the matrix
```{r}
skillsNeededMatrix <- matrix(as.numeric(skillsNeeded[,2:(nSkills+1)]),
                             nrow = nPeople,
                             ncol = nSkills, 
                             dimnames = list(namesOfPeople,namesOfSkills))

skillsNeededMatrix
```

2. Now, let's tackle skillsOffered:

# Declare all the  necessary variables
```{r}
skillsOffered <- as.matrix(read_excel("skillsOffered.xlsx"))
namesOfPrograms <- skillsOffered[,1]
nPrograms <- nrow(skillsOffered)

skillsOffered
```

# Build the matrix
```{r}
skillsOfferedMatrix <- matrix(as.numeric(skillsOffered[,2:(nSkills+1)]),
                             nrow=nPrograms,ncol=nSkills,
                             dimnames = list(namesOfPrograms,namesOfSkills))
skillsOfferedMatrix
```

3. Let's read in the programConflicts:

# Declare all the  necessary variable
```{r}
programConflicts <- as.matrix(read_excel("programConflicts.xlsx"))
programConflicts
```

Note: The conflicting programs are locked up in a text string. To get them out, we need to split the string and force it into numeric format that we will use later.

4. Now, let's create our global variables

```{r}
namesOfVariables <- c()#initialize an empty list

for(person in namesOfPeople){
  for(program in namesOfPrograms){
   newVariable <- paste("x",person,program,sep = ".") 
    namesOfVariables <- c(namesOfVariables,newVariable)#tack the new one onto the end of the list
  }
}

namesOfVariables
nVariables <- length(namesOfVariables)
nVariables
```

5. Get the objective vector, and multiply the vector as per the number of people

# Declare all the  necessary variabless
```{r}
programCostIndex <- grep("Program Cost",colnames(skillsOffered))
programCost <- (skillsOffered[,programCostIndex])
objective_vector_row <- c(rep(as.numeric(programCost), nPeople))
objective_vector_row
```

# Build the matrix
```{r}
objectiveFunction <- matrix(objective_vector_row, nrow = 1)
colnames(objectiveFunction) <- namesOfVariables
objectiveFunction
```


## Constraint Matrix Set Up
   - - - - - - - - - - - - 
   
- Let's set up our constraint matrix. Our strategy here will be to create an empty matrix and add to it row by row using `rbind()`.


1. Create the empty matrix

```{r}
constraintMatrix <- matrix(0, nrow = 0, ncol = nVariables,
                           dimnames = list(c(),namesOfVariables))
inequalities <- vector("character", length = 0)
rhs <- vector("numeric", length = 0)
```

2. Let's create the first set of constraints that say that at most 3 employees can attend the same program.

    I. Make the first constraint
   
```{r}
zeroRow <- rep(0, nVariables)

for (program in namesOfPrograms) {
  pattern <- paste0(program, "$")
  indexToChange <- grep(pattern, namesOfVariables)
  newConstraint <- zeroRow
  
  newConstraint[indexToChange] <- 1
  constraintMatrix <- rbind(constraintMatrix,newConstraint)
}

rownames(constraintMatrix) <- NULL
firstConstraintRow <- nrow(constraintMatrix)
firstConstraintRow
ncol(constraintMatrix)
options(max.print = 10000)
constraintMatrix
```

# Update the inequalities 
```{r}
inequalities <- c(inequalities, rep("<=", firstConstraintRow))
inequalities
```

# Update the rhs 
```{r}
maxEmp <- 3
rhs <- c(rhs, rep(maxEmp, firstConstraintRow))
rhs
```


    II. Let's create the second set of constraints that say each person is trained to the skills they need.

```{r}
zeroRow <- rep(0, nVariables)

# double loop to go through every person's required skill
for (person in namesOfPeople) {
  for (skill in namesOfSkills) {
    
    # check if the person needs that skill
    if ( skillsNeededMatrix[person, skill] == 1 ) {
       
       indexToChange <- c()
      
      # loop through the vertical columns of the skills required to see which programs offer it
      for (program in namesOfPrograms) {
        
        # check to see if a program offers that skill
        if ( skillsOfferedMatrix[program, skill] == 1) {
          
            pattern <- paste0(person, ".", program, "$")
            matching_index <- grep(pattern, namesOfVariables)
            indexToChange <- c(indexToChange, matching_index)
        }
      }
      
           newConstraint <- zeroRow
           newConstraint[indexToChange] <- 1
           constraintMatrix <- rbind(constraintMatrix,newConstraint)
            
    }
  }
}


rownames(constraintMatrix) <- NULL
secondConstraintRow <- nrow(constraintMatrix) - firstConstraintRow
secondConstraintRow
nrow(constraintMatrix)
ncol(constraintMatrix)
options(max.print = 10000)

# only print the new additions to the constraint Matrix
constraintMatrix[(firstConstraintRow + 1):nrow(constraintMatrix), ]
```

# Update the inequalities 
```{r}
inequalities <- c(inequalities, rep(">=", secondConstraintRow))
inequalities
```

# Update the rhs 
```{r}
maxVal <- 1
rhs <- c(rhs, rep(maxVal, secondConstraintRow))
rhs
```


    III. Let's create the third set of constraints that say that no one can attend two training sessions that conflict with one another.

```{r}
zeroRow <- rep(0, nVariables)

# assign M to the number of Programs
M <- nPrograms

# double loop to assume everybody considers a program
for ( person in namesOfPeople) {
  for ( program in namesOfPrograms) {
    
   # get a list of conflicting programs for program p
   p <- grep( paste0(program, "$"), namesOfPrograms)
   conflictingPrograms <- as.numeric(unlist(strsplit(programConflicts[p, 2],split = ", ")))
  
   indexToChange <- c()
   
   for (conflictingProgram in conflictingPrograms) {
       pattern <- paste0(person, ".", namesOfPrograms[conflictingProgram], "$")
       matching_index <- grep(pattern, namesOfVariables)
       indexToChange <- c(indexToChange, matching_index)
   }
   
   
   newConstraint <- zeroRow
   newConstraint[ grep( paste0(person, ".", program, "$") ,namesOfVariables)  ] <- M
   newConstraint[indexToChange] <- 1
   constraintMatrix <- rbind(constraintMatrix,newConstraint)
    
  }
}


rownames(constraintMatrix) <- NULL
thirdConstraintRow <- nrow(constraintMatrix) - secondConstraintRow - firstConstraintRow
thirdConstraintRow
nrow(constraintMatrix)
ncol(constraintMatrix)
options(max.print = 10000)

# only print the new additions to the constraint Matrix
constraintMatrix[(secondConstraintRow+firstConstraintRow + 1):nrow(constraintMatrix), ]

```

# Update the inequalities 
```{r}
inequalities <- c(inequalities, rep("<=", thirdConstraintRow))
inequalities
```

# Update the rhs 
```{r}
rhs <- c(rhs, rep(M, thirdConstraintRow))
rhs
```


## Let's solve it in R
   - - - - - - - - - -

 - We are looking for the optimal minimum cost and who should go to which program to obtain that cost.
   Note: Set `all.bin=TRUE` as an option in `lp()` to set all decision variables as binary.

# Lp Solve
```{r}
lp_solution <- lp(direction = "min",
                  const.mat = constraintMatrix,
                  const.dir = inequalities,
                  const.rhs = rhs,
                  objective.in = objectiveFunction,
                  all.bin = TRUE)

lp_solution$status
lp_solution$solution
lp_solution$objval
lp_solution$solution==1
namesOfVariables[lp_solution$solution==1]

```



